Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm} Stagnating State Funding for Higher Education and its Effect on Faculty

### 2024-09-17 text/paper.tex
Made some edits.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
        For helpful comments I thank
        Levon Barseghyan,
        Francine Blau,
        Romain Campbell,
        Maddalena Davoli,
        Ryan Dycus,
        Ronald Ehrenberg,
        Michael Lovenheim,
        Jake Meyer,
        Douglas Miller,
        and Evan Riehl.
        I thank seminar participants at Cornell University (2023, 2024), and the Eastern Economic Association Annual Meetings (2024) for helpful discussion.
        This project's repository, including all materials and anonymised data needed for replication, is publicly available at 
        \url{https://github.com/shoganhennessy/state-funding-faculty}.
        Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.
    } \\
    \vspace{0.1cm}
    Economics Department, Cornell University\footnote{
        Address: Uris Hall \#447, Economics Department, Cornell University NY 14853 USA.
    }
}
% Title
\title{Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm}
    Stagnating State Funding for Higher Education and its Effect on Faculty}
\date{This version: \today \\ \vspace{0.1cm}
    \href{https://github.com/shoganhennessy/state-funding-faculty/blob/main/state-funding-faculty-2024.pdf}{Newest version available here.}
}
\rhead{\today}
\lhead{Less Funding, More Lecturers, and Fewer Professors.}
% Begin
\begin{document}
\maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
    \noindent
    \input{sections/00-abstract.tex}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Starting the paper
\newpage
\setcounter{page}{1}
\doublespacing
% Introduction section
\noindent
\input{sections/01-introduction.tex}
% Data section
\input{sections/02-data.tex}
% Conceptual Framework
\input{sections/03-concepts.tex}
% Empirical methods
\input{sections/04-empirical.tex}
% Empirical results
\input{sections/05-results.tex}
% Dicussion
%\input{sections/06-discussion.tex}
% Conclusion
\input{sections/07-conclusion.tex}
% Rev of Econ Ed Statements.
\subsubsection*{Data Statement}
Every source of data, and the code to fully replicate the entire analysis is freely available at the following link:
\url{https://github.com/shoganhennessy/state-funding-faculty}.
IBHED data, while public knowledge, include personally identifiable names, so the data included in the replication package have names replaced with non-identifiable placeholders.
\subsubsection*{Credit Author Statement}
As the only author on this paper, I am responsible for all contributions in this paper.
\subsubsection*{Declaration of Competing Interest}
I have no competing interests to declare.
% Separated Figures \& Tables
\input{sections/09-figures-tables.tex}
% Bibliography
\singlespacing
\bibliographystyle{agsm}
\bibliography{sections/08-bibliography.bib}
% Appendix
\input{sections/10-appendix.tex}
\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
        For helpful comments I thank
        Levon Barseghyan,
        Francine Blau,
        Romain Campbell,
        Maddalena Davoli,
        Ryan Dycus,
        Ronald Ehrenberg,
        Michael Lovenheim,
        Jake Meyer,
        Douglas Miller,
        and Evan Riehl.
        I thank seminar participants at Cornell University (2023, 2024), and the Eastern Economic Association Annual Meetings (2024) for helpful discussion.
        This project's repository, including all materials and anonymised data needed for replication, is publicly available at 
        \url{https://doi.org/10.5281/zenodo.11373226}.
        Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the \href{https://github.com/shoganhennessy/state-funding-faculty/}{Github project}.
    } \\
    \vspace{0.1cm}
    Economics Department, Cornell University\footnote{
        Address: Uris Hall \#447, Economics Department, Cornell University NY 14853 USA.
    }
}
% Title
\title{Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm}
    Stagnating State Funding for Higher Education and its Effect on Faculty}
\date{First version: 1 September 2022 \\ \vspace{0.15cm}
    This version: \today \\ \vspace{0.15cm}
    \href{https://github.com/shoganhennessy/state-funding-faculty/blob/main/state-funding-faculty-2024.pdf}{Newest version available here.}
    \\ \vspace{-0.5cm}
}
\rhead{\today}
\lhead{Less Funding, More Lecturers, and Fewer Professors.}
% Begin
\begin{document}
\maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
    \noindent
    \input{sections/00-abstract.tex}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Starting the paper
\newpage
\setcounter{page}{1}
\doublespacing
% Introduction section
\noindent
\input{sections/01-introduction.tex}
% Data section
\input{sections/02-data.tex}
% Conceptual Framework
\input{sections/03-concepts.tex}
% Empirical methods
\input{sections/04-empirical.tex}
% Empirical results
\input{sections/05-results.tex}
% Dicussion
%\input{sections/06-discussion.tex}
% Conclusion
\input{sections/07-conclusion.tex}
% Rev of Econ Ed Statements.
\subsubsection*{Credit Author Statement}
As the only author on this paper, I am responsible for all contributions in this paper.

\subsubsection*{Data Statement}
Every source of data, and the code to fully replicate the entire analysis is freely available at the following link:
\url{https://doi.org/10.5281/zenodo.11373226}.

IBHED data, while public knowledge, include personally identifiable names, so the data included in the replication package have names replaced with non-identifiable placeholders.

\subsubsection*{Declaration of Competing Interest}
I have no competing interests to declare.

This research did not receive any specific grant from funding agencies in the public, commercial, or not-for-profit sectors.

% Separated Figures \& Tables
\input{sections/09-figures-tables.tex}
% Bibliography
\singlespacing
\bibliographystyle{agsm}
\bibliography{sections/08-bibliography.bib}
% Appendix
\input{sections/10-appendix.tex}
\end{document}

### 2024-09-17 text/sections/01-introduction.tex
Made some edits.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction section
%\section{Introduction}
%\label{sec:intro}
States fund their public universities with roughly the same amount today as they did in the 1990s, despite enrolment rising by over 50\%.
At the same time, public universities have systematically substituted away from professors towards lecturers, who often work on contingent and lower-salaried contracts.
Previous research shows state funding cuts led to worsening student outcomes \citep{NBERw23736,NBERw27885}, but there is little evidence on whether these funding cuts impacted faculty.
In this paper, I show that falling state funding caused a change in the composition of faculty at US public universities, substituting professors for more lecturers.

per-student state funding for higher education has stagnated the last thirty years.
The average public university received around \$11,600 of state funding per-student in 1990 and \$8,300 in 2021, while the number of lecturers per-student increased by 109\% and professors per-student decreased by $23$\% over 1990--2021.
I use a shift-share instrumental variables approach to estimate how faculty are affected by these funding cuts.
Universities employ $4.4$\% more lecturers per-student following a $10$\% funding cut, and correspondingly rely less on tenure-track faculty by employing $1.4$\% fewer tenure-track professors and $1.2$\% fewer tenured professors per-student.
In a secondary analysis of all public university faculty in the state of Illinois, I show that incumbent faculty were unaffected, in terms of wages, promotion and quit rates.
This implies that the faculty substitution arose by disrupting hiring of new tenure-track/tenured professors, which is supported supplementary data on professor hiring at US public universities over 2010--2021.
This paper shows the institutional effects of stagnating state funding for higher education, and offers one possible mechanism for how funding cuts affected student outcomes.

@@

Mine is the first paper to provide causal evidence on the impact of state funding on faculty composition and wages.
The closest related research on higher education funding examined how university spending affected graduation rates and levels of student debt \citep{NBERw23736,NBERw27885}, and university finances \citep{miller2022making,bound2019public,brown2014endowment}.
For faculty outcomes, descriptive work measures how US and Canadian faculty salaries are correlated with other factors \citep{bratsberg2003negative,martinello2009faculty,pfeffer1992salary,hilmer2020labor}, and applied theory work models university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
Other papers measure university decision--making after endowment shocks \citep{brown2014endowment}, and trends in the academic job market following the 2008 recession \citep{turner2014impact}.
These papers measure changes in state funding and university revenues, but do not measure effects of state funding cuts on faculty composition or salaries, or other outcomes related to university instruction.
My results provide evidence that the long term trends in higher education funding and faculty outcomes are causally related, contributing to the literature on faculty outcomes \citep{ehrenberg2003studying} and trends in US higher education and funding \citep{hoxby2009changing,ehrenberg2012american}.

Another contribution of my paper is to provide evidence on possible mechanisms connecting state funding cuts for higher education and worsening student outcomes.
In 2021, around 65\% of lecturers at US public universities are employed part-time (IPEDS); this arrangement is often arranged as an adjunct contract, where faculty are paid based on how many courses they teach, without employment benefits \citep{aau2014Characteristics}.
While funding cuts force universities to substitute towards lecturers, there is consistent evidence that students are worse off being taught by part-time lecturers than full-time faculty.
Universities that rely on more lecturers have lower student retention and graduation rates \citep{jaeger2011examining,ehrenberg2005tenured}.
Similarly, students at a large public university have consistently worse education outcomes if taught by adjunct lecturers than those taught by full-time faculty, largely thanks to the worse working conditions adjunct lecturers face \citep{zhu2021limited}.\footnote{
    Alternatively, \cite{bettinger2010does,figlio2015tenure} study lecturers at a highly selective universities, finding that adjunct lecturers improve subject retention for undergraduates compared to professors.
    This paper studies faculty at public universities, which are by-majority not selective, and where lecturers are more often employed part-time with less institutional support.
}
Combining the evidence that adjunct lecturers lead to worse student outcomes with the trends in education finances and performance \citep{NBERw23736,NBERw27885}, implies that substituting towards lecturers contributed to deteriorating student outcomes at public universities.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction section
\section{Introduction}
\label{sec:intro}
States fund their public universities with roughly the same amount today as they did in the 1990s, despite enrolment rising by over 50\%.
At the same time, public universities have systematically substituted away from professors towards lecturers, who often work on contingent and lower-salaried contracts.
Previous research shows state funding cuts led to worsening student outcomes \citep{NBERw23736,NBERw27885}, but there is little evidence on whether these funding cuts impacted faculty.
In this paper, I show that falling state funding caused a change in the composition of faculty at US public universities, substituting professors for more lecturers.

Per-student state funding for higher education has stagnated the last thirty years.
The average public university received around \$11,600 of state funding per-student in 1990 and \$8,300 in 2021, while the number of lecturers per-student increased by 109\% and professors per-student decreased by $23$\% over 1990--2021.
I use a shift-share instrumental variables approach to estimate how faculty are affected by these funding cuts.
Universities employ $4.4$\% more lecturers per-student following a $10$\% funding cut, and correspondingly rely less on tenure-track faculty by employing $1.4$\% fewer tenure-track professors and $1.2$\% fewer tenured professors per-student.
In a secondary analysis of all public university faculty in the state of Illinois, I show that incumbent faculty were unaffected, in terms of wages, promotion and quit rates.
This implies that the faculty substitution arose by disrupting hiring of new tenure-track/tenured professors, which is supported supplementary data on professor hiring at US public universities over 2010--2021.
This paper shows the institutional effects of stagnating state funding for higher education, and offers one possible mechanism for how funding cuts affected student outcomes.

@@

Mine is the first paper to provide causal evidence on the impact of state funding on faculty composition and wages.
The closest related research on higher education funding examined how university spending affected graduation rates and levels of student debt \citep{NBERw23736,NBERw27885}, and university finances \citep{miller2022making,bound2019public,brown2014endowment}.
For faculty outcomes, descriptive work measures how US faculty salaries are correlated with other factors \citep{hilmer2020labor}, and applied theory work models university decision-making and faculty hiring (see e.g., \citealt{abe2015implications,johnson2009jep,NBERc13879}).
Other papers measure university decision--making after endowment shocks \citep{brown2014endowment}, and trends in the academic job market following the 2008 recession \citep{turner2014impact}.
These papers measure changes in state funding and university revenues, but do not measure effects of state funding cuts on faculty composition or salaries, or other outcomes related to university instruction.
My results provide evidence that the long term trends in higher education funding and faculty outcomes are causally related, contributing to the literature on faculty outcomes \citep{ehrenberg2003studying} and trends in US higher education and funding \citep{hoxby2009changing,ehrenberg2012american}.

Another contribution of my paper is to provide evidence on possible mechanisms connecting state funding cuts for higher education and worsening student outcomes.
In 2021, around 65\% of lecturers at US public universities are employed part-time (IPEDS); this arrangement is often arranged as an adjunct contract, where faculty are paid based on how many courses they teach, without employment benefits \citep{aau2014Characteristics}.
While funding cuts force universities to substitute towards lecturers, there is consistent evidence that students are worse off being taught by part-time lecturers than full-time faculty.
Universities that rely on more lecturers have lower student retention and graduation rates \citep{jaeger2011examining,ehrenberg2005tenured}.
Similarly, students at a large public university have consistently worse education outcomes if taught by adjunct lecturers than those taught by full-time faculty, largely thanks to the worse working conditions adjunct lecturers face \citep{zhu2021limited}.
Alternatively, \cite{bettinger2010does,figlio2015tenure} study lecturers at a highly selective universities, finding that adjunct lecturers improve subject retention for undergraduates compared to professors.
This paper studies faculty at public universities, which are by-majority not selective, and where lecturers are more often employed part-time with less institutional support.
Combining the evidence that adjunct lecturers lead to worse student outcomes with the trends in education finances and performance \citep{NBERw23736,NBERw27885}, implies that substituting towards lecturers contributed to deteriorating student outcomes at public universities.
